---
title: "Complex Sample Survey Data Analysis: NSDUH 2015-2019"
author: "F. Hunter McGuire, MPH"
date: "February 28, 2022"
output: 
  rmarkdown::html_document:
    theme: cosmo
    highlight: pygments
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

## **Background and Purpose**

The purpose of this tutorial is to provide brief introduction to conducting design-based analysis of complex sample survey data using the `survey` package in `R`.

We will be using data from the **National Survey on Drug Use and Health (NSDUH)**. NSDUH is a repeated cross-sectional household survey distributed annually to a sample of non-institutionalized residents of the United States. Its primary aim is to monitor drug and alcohol use by providing nationally-representative estimates of substance use behaviors, mental health status, and correlates of these health outcomes/behaviors. Additional information on NSDUH can be found [here](https://nsduhweb.rti.org/respweb/about_nsduh.html).

To improve estimate precision and reduce overall survey costs, households (and specific individuals within households) who participate in NSDUH are selected through a complex set of **cluster and stratified sampling** procedures. Certain population subgroups (including younger people, Hispanic people, Black/African-American people) are also **oversampled relative to their general population size** to produce more reliable estimates (i.e., through standard error reduction). This ultimately generates a sample in which there is **unequal probability of selection** across sampling units, and produces somewhat homogeneous samples within clusters due to shared community characteristics (i.e., **nonindependence**). Together, these violate simple random sample assumptions (equal probability of selection, independence of observations) common across basic statistical procedures. 

**In order to generate accurate nationally-representative estimates (e.g., proportions, means, standard errors) from complex sample survey data, we have to conduct design-based analysis** (i.e., analysis that adjusts for unequal probability of selection). For more information on NSDUH sample design can be found in the [NSDUH 2019 Sample Design Report](https://www.samhsa.gov/data/sites/default/files/reports/rpt34664/NSDUHmrbSampleDesign2019.pdf). Guidance for statistical analysis can be found in the [NSDUH 2019 Statistical Inference Report](https://www.samhsa.gov/data/sites/default/files/reports/rpt34666/NSDUHmrbStatInference2019.pdf).

### Data frame information: NSDUH 2015-2019

The data frame included in this tutorial (`nsduh_20152019_subset.RData`) includes 5 combined waves (2015, 2016, 2017, 2018, 2019) and a subset of 15 variables to reduce the overall file size.

**Combining waves**: To increase the sample size of relatively small population subgroups, it is common practice in public health research to combine multiple waves from national cross-sectional datasets like NSDUH. However, additional procedures are needed to adjust sampling weights, and these simple steps are outlined in this tutorial.

**Unconditional analysis of a subpopulation**: In addition, researchers are often interested in studying a specific subpopulation (e.g., adolescents). Here, it is neccessary to identify the subpopulation within the overall dataset for the analysis at hand. **Avoid dropping unused observations as this removes key information used to generate variance estimates.** Additional background information on subpopulation analysis can be found in [West, Berglund, & Herringa, 2008](https://journals.sagepub.com/doi/pdf/10.1177/1536867X0800800404) and in [Lumley, 2021](https://cran.r-project.org/web/packages/survey/vignettes/domain.pdf).

### Key issues and procedures covered
1. Load the data frame and identify a subpopulation of interest
2. Setup design-based analysis using `svydesign`
3. Descriptive statistics using `svymean`, `svyby`, and `svyciprop`
4. T-tests and design-based Wald (chi-square) tests of independence
5. Regression analysis using `svyglm`
6. Regression diagnositics using the `svydiags` package (under construction!)

Note: A useful resource for analyzing complex sample data in R is available [here](http://asdfree.com/national-study-on-drug-use-and-health-nsduh.html).

## **Setup**

First, we need to load the NSDUH 2015-2019 dataset and the necessary `R` packages relevant for this analysis.

The most important package is the `survey` package [(Lumley, 2021)](https://cran.r-project.org/web/packages/survey/survey.pdf). This package provides flexible tools to generate design-based estimates that account for the complex sample survey design of NSDUH.

We will also use the `svydiags` package to compute diagnostic data for regression models fitted with complex survey data [Valliant, 2018](https://cran.r-project.org/web/packages/svydiags/svydiags.pdf)

This project also uses the core suite of `tidyverse` packages to assist with data management procedures [(Wickham, 2021)](https://cran.r-project.org/web/packages/tidyverse/tidyverse.pdf), the `stats` package to run statistical procedures [(R Core Team)](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/00Index.html), and the `jtools` package for miscellaneous statistical procedures [(Long, 2021)](https://cran.r-project.org/web/packages/jtools/jtools.pdf).

```{r setup, message=FALSE}
# set working directory (update file path to where you stored the dataset)
setwd("C:/Users/Hunter McGuire/Desktop/Box Sync/Invited Talks and Guest Lectures/Complex Sample Survey Analysis/")

# load packages
library(survey) ## run design-based analysis
library(svydiags) ## regression diagnostics
library(tidyverse) ## data management
library(stats) ## for running statistical procedures
library(jtools) ## Analysis and Presentation of Social Scientific Data
```

## **Load the data frame and identify a subpopulation of interest**

The following code outlines some initial data management steps. I have renamed some variables to make it easier to quickly identify and interpret results. See the [NSDUH 2019 Public Use File Codebook](https://www.datafiles.samhsa.gov/sites/default/files/field-uploads-protected/studies/NSDUH-2019/NSDUH-2019-datasets/NSDUH-2019-DS0001/NSDUH-2019-DS0001-info/NSDUH-2019-DS0001-info-codebook.pdf) for more information on variables and coding schemes in NSDUH.

**Subpopulation**: In this tutorial, the subpopulation of interest are adults (aged 18+) in the United States with non-missing data on the following variables: depression_bin (past-year major depressive episode, binary coded), stress_cont (Kessler K6 distress scale, continuous: range 0-24), race/ethnicity (newrace2), gender (irsex), sexident (sexual orientation), annual household income level (income), and anyhlti2 (any health insurance).

Following the procedures below, we identify a subpopulation of N=207,479 participants who meet our study eligibility criteria.

```{r datamgmt, message=FALSE}
# Load the data set
load(file="nsduh_20152019_subset.RData")
nsduh <- rename(nsduh_20152019_subset) # shorten the data set name
rm(nsduh_20152019_subset) # remove old data frame

# set invalid responses to missing (NA)
library(naniar)
nsduh <- nsduh %>% 
  replace_with_na(replace = list(sexident = c(85, 89, 94, 97, 98, 99))) %>% 
  replace_with_na(replace = list(anyhlti2 = c(94, 97, 98)))

# create factor variables for categorical vars
nsduh$amdeyr <- as.factor(nsduh$amdeyr) # past-year major depressive episode (MDE)
nsduh$newrace2 <- as.factor(nsduh$newrace2) # racial/ethnic identity
nsduh$irsex <- as.factor(nsduh$irsex) # gender (binary woman/man)
nsduh$sexident <- as.factor(nsduh$sexident) # sexual orientation
nsduh$catag6 <- as.factor(nsduh$catag6) # age category
nsduh$income <- as.factor(nsduh$income) # annual household income level
nsduh$anyhlti2 <- as.factor(nsduh$anyhlti2) # any health insurance (1=yes)
nsduh$year <- as.factor(nsduh$year) # year of NSDUH survey completion

# rename/mutate variables for ease of use in analysis
nsduh <- nsduh %>%
  mutate(
    white = factor(ifelse(newrace2 == "1", 1, 0)), # White
    black = factor(ifelse(newrace2 == "2", 1, 0)), # Black/African-American
    naan = factor(ifelse(newrace2 == "3", 1, 0)), # Native American or Alaska Native
    nhpi = factor(ifelse(newrace2 == "4", 1, 0)), # Native Hawaiian/Pacific Islander
    asian = factor(ifelse(newrace2 == "5", 1, 0)), # Asian
    multi = factor(ifelse(newrace2 == "6", 1, 0)), # Multiracial
    hispanic = factor(ifelse(newrace2 == "7", 1, 0)), # Hispanic
    man = factor(ifelse(irsex == "1", 1, 0)), # Men
    woman = factor(ifelse(irsex == "2", 1, 0)), # Women
    hetero = factor(ifelse(sexident == "1", 1, 0)), # Heterosexual
    gay = factor(ifelse(sexident == "2", 1, 0)), # Gay
    bi = factor(ifelse(sexident == "3", 1, 0)), # Bisexual
    depression_bin = factor(ifelse(amdeyr == "1", 1, 0)), # Past-year MDE
    healthins = factor(ifelse(anyhlti2 == "1", 1, 0)), # any health insurance
    stress_cont = k6scmon) %>% # K6 Distress Scale
  mutate(
    subpop = factor(ifelse(!is.na(newrace2) & !is.na(sexident) & !is.na(irsex) & !is.na(depression_bin) & !is.na(stress_cont) & !is.na(income) & !is.na(catag6) & !is.na(healthins) & (catag6!="1"), 1, 0))) # this creates a "subpop" variable to identify all observations in the dataset that have non-missing values on each of the study variables of interest.

table(nsduh$subpop) # This indicates that 207,479 participants have non-missing data on our study variables.
```


## **Setup design-based analysis**

NSDUH includes the following survey weight variables: **vestr** (variance estimation stratum), **verep** (variance estimation cluster replicates), and **analwt_c** (person-level analysis weight). The `svydesign` procedure takes these variables as arguments to generate a design object that we can easily include in our analysis going forward.

**Combining waves**: Since we are combining multiple data collection waves in this tutorial, we need to divide the person-level analysis weight by the number of waves included.

**Subpopulation**: Using the "subpop" variable we generated in the prior step, we can simply subset the `svydesign` object to includ our subpopulation (N=207,479).

```{r svydesign, message=FALSE}
# divide person-level analysis weight by 5 (5 waves included: 2015-2019)
nsduh$adjwt_5 <- nsduh$analwt_c/5
# Note: if only using one wave of data (e.g., only 2015), the above step is not required.

# build up the primary survey object
nsduh_design <- 
  svydesign(
    id = ~verep,
    strata = ~vestr,
    weights = ~adjwt_5,
    data = nsduh,
    nest = TRUE)

# svydesign for the subpopulation -- we will use this throughout
nsduh_design_sp <- subset(
  nsduh_design, 
  subpop=="1")
summary(nsduh_design_sp) ## describes the svydesign object
```

## **Descriptive statistics**

```{r descriptives, message=FALSE}
# Survey-weighted means and 95% CI
## Estimate mean K6 Distress Scale scores for entire subpop
k6mean <- svymean(~stress_cont, nsduh_design_sp)
k6mean
## Estimate 95% confidence interval for entire subpop
confint(k6mean)
## Estimate mean for individual subgroups (gender)
k6mean_gender <- svyby(~stress_cont, ~woman, design=nsduh_design_sp, svymean)
k6mean_gender
## Estimate 95% confidence interval for individual subgroups
confint(k6mean_gender)

# Survey-weighted proportions and 95% CI
## Race/ethnicity
svyciprop(~white, nsduh_design_sp, method="logit")
svyciprop(~black, nsduh_design_sp, method="logit")
svyciprop(~hispanic, nsduh_design_sp, method="logit")
svyciprop(~asian, nsduh_design_sp, method="logit")
svyciprop(~naan, nsduh_design_sp, method="logit")
svyciprop(~nhpi, nsduh_design_sp, method="logit")
svyciprop(~multi, nsduh_design_sp, method="logit")
## Gender
svyciprop(~woman, nsduh_design_sp, method="logit")
svyciprop(~man, nsduh_design_sp, method="logit")
##Sexual orientation
svyciprop(~hetero, nsduh_design_sp, method="logit")
svyciprop(~gay, nsduh_design_sp, method="logit")
svyciprop(~bi, nsduh_design_sp, method="logit")
## Past-year major depressive episode
svyciprop(~depression_bin, nsduh_design_sp, method="logit")
## Any health insurance
svyciprop(~healthins, nsduh_design_sp, method="logit")
## Age category
svyciprop(~I(catag6==2), nsduh_design_sp, method="logit")
svyciprop(~I(catag6==3), nsduh_design_sp, method="logit")
svyciprop(~I(catag6==4), nsduh_design_sp, method="logit")
svyciprop(~I(catag6==5), nsduh_design_sp, method="logit")
svyciprop(~I(catag6==6), nsduh_design_sp, method="logit")
## Income level
svyciprop(~I(income==1), nsduh_design_sp, method="logit")
svyciprop(~I(income==2), nsduh_design_sp, method="logit")
svyciprop(~I(income==3), nsduh_design_sp, method="logit")
svyciprop(~I(income==4), nsduh_design_sp, method="logit")

## Graphing continuous variables
# Note: Currently ggplot2 is not equipped to handle complex sample data, but there are some packages online that are beginning to overcome this hurdle.
svyhist(~stress_cont, nsduh_design_sp) # positive skew
svyboxplot(~stress_cont~1, nsduh_design_sp, all.outliers = TRUE) 
```

## **T-tests and design-based Wald (chi-square) tests of independence**

```{r analysis_1, message=FALSE}
# Do Kessler-6 Distress Scale scores differ between men and women?
## Independent samples t-test: svyttest(cont~cat, design)
t <- svyttest(stress_cont ~ woman, 
              design=nsduh_design_sp)
t 
## Yes, women have higher scores.
## Mean difference = 0.61 (95% CI: 0.56, 0.67)

## Notes: 
## paired sample t-test can be run as: svyttest(I(var1-var2)~0, design)
## one-sample t-test can be run as: svyttest(var~0, design)

## Wilcoxon signed rank test (non-parametric independent t-test)
w <- svyranktest(stress_cont ~ woman, nsduh_design_sp, test=c("wilcoxon"))
w


# Does the depression prevalence differ by sexual orientation?
## Design-based Wald (chi-square) test of association: svychisq(~cat+cat, design, statistic="adjWald")
c <- svychisq(~depression_bin + sexident,
              design=nsduh_design_sp, 
              statistic="adjWald")
c
## Yes, there are differences in depression by sexual orientation.
```

## **Regression analysis**

```{r analysis_2, message=FALSE}
# Linear regression
## Are there differences in Kessler-6 Distress Scale scores by race/ethnicity?
### Simple linear regression
a1 <- svyglm(stress_cont ~ black + hispanic + asian + naan + nhpi + multi, 
             nsduh_design_sp, 
             family=gaussian())
summary(a1)
# Yes, compared to white participants: Black, Native Hawaiian/Pacific Islander, and Hispanic participants had lower stress scores; Multiracial participants had higher stress scores.

## Is race/ethnicity still a significant predictor of Kessler-6 Distress Scale scores after controlling for income level, age category, and health insurance status?
### Multivariate linear regression
a2 <- svyglm(stress_cont ~ black + hispanic + asian + naan + nhpi + multi + income + catag6 + anyhlti2,
             nsduh_design_sp, 
             family=gaussian())
summary(a2) ## Note: we should assess regression diagnostics before reporting results

### Logistic regression
a3 <- svyglm(depression_bin ~ gay + bi,
             nsduh_design_sp,
             family=quasibinomial())
summary(a3)
```

## **Regression diagnostics for complex sample data**

```{r diagnostics, message=FALSE}
# Coming soon!

# Linear regression diagnostics (`svydiags` package)
## Linearity between X variables and Y
## Multicollinearity
## Homoscedasticity of residual variance
## Normality of residuals

# Logistic regression assumptions
## X variables are linearly related to the logit of the Y variable
```

